Investigating the Impact of Macroeconomic Indicators on the FTSE 250 Index Using Machine Learning Methods

### 2024-08-14 tex/paper.tex
lit review

The Fisher hypothesis states that nominal interest rates adjust to expected 
inflation, implying market efficiency as firms correctly anticipate inflation 
and adjust prices. This was tested this in the US stock market by \citeboth{jaffe1976}, and
\citeboth{bodie1976} who both found a negative relationship between inflation and stock returns.
To account for this,
\citeboth{mogdiliani1979} proposed the Inflation Illusion Hypothesis, 
arguing that investors overestimate the discount rate in valuations, leading to mispricing. 
\citeboth{gultekin1983} and \citeboth{firth1979}, however, found that the relationship between nominal stock returns and inflation in the UK is positive, and that real returns on UK stocks have remained stable even as inflation varied.
Later, \citeboth{hasan2008} provided evidence against this, challenging the Fisher hypothesis.

@@

Interest rates, being the cost of borrowing, directly impact firms by increasing financing costs, reducing profitability, and lowering growth prospects. For consumers, higher rates raise borrowing costs, leading to reduced spending and demand, which further lowers corporate profitability. 
The empirical research supports this theory: \citeboth{alam2009} found a significant inverse relationship between interest rates and stock prices across multiple countries from 1988 to 2003; 
\citeboth{talla2013} observed a negative relationship between interest rates and stock prices on the Stockholm Stock Exchange; 
\citeboth{demir2019} reported similar findings for the Turkish Stock Exchange. In the case of the UK, \citeboth{neifar2023} used a Non-Linear ARDL model to find a significantly negative long-run relationship between interest rates and the UK stock market. Higher interest rates also encourage a shift from stocks to safer 
assets like bonds, this is known as the flight-to-quality phenomenon, where rising bond yields make bonds more attractive compared to stocks. \citeboth{asgharian2016} provided evidence for this, particularly in developed markets like the UK. This typically results in a negative correlation between stock prices and interest rates.

@@

There is a large body of literature regarding time-series forecasting. First, 
a discussion on the feasibility of such forecasting is presented: the Efficient 
Market Hypothesis. Following this, machine learning methods for forecasting are 
presented,

@@

The consistent success of various high-profile investors in the decades 
after the development of the EMH have challenged 
its central argument, something acknowledged in the revised formulation of the EMH 
in \citeboth{fama1991}. Furthermore, the empirical research is not in a
greement on whether the London Stock Exchange is efficient at any of the three levels. 
\citeboth{libberton2010} and \citeboth{rounaghi} give evidence of weak-form efficiency,
however, \citeboth{borges2010}, \citeboth{asghar2023} and \citeboth{bhavsar2015} al provide evidence to reject the EMH. 
Some studies such as \citeboth{rosini2020} even found that the 
efficiency of the FTSE100 and FTSE250 varied over time. The empirical findings then suggest 
that it is not impossible to determine movements in the 
price of stock indices on the FTSE250 using macroeconomic variables.

@@

Before constructing the ARDL model a natural logarithmic transform is applied
on the variables, this is done because in a log-transformed model,
the coefficients can be interpreted as elasticities. Put simply, 
a coefficient in a model, where both the dependent and independent variables
are in logarithmic form, represents the percentage change in the 
dependent variable for a one percent change in an independent variable, 
which furnishes us with easily interpretable results. Having done this,
the level at which the variables are stationary\footnote{Stationarity refers to the statistical property of a time-series where 
its mean, variance, and autocorrelation structure do not change over time. Either a variable is stationary at level ($I(0)$), requiring no alteration,
or it is stationary after taking the first difference ($I(1)$). 
Non-stationarity leads to spurious regression where relationships that 
are not actually present between two time-series are produced.} is checked.
Various techniques exist for stationarity testing, however in this study the 
Augmented Dickey-Fuller (ADF) unit root test, was used.

>>>>

The Fisher hypothesis states that nominal interest rates adjust to expected 
inflation, implying that markets are efficient as firms correctly anticipate inflation 
and adjust prices. This was tested this in the US stock market by \citeboth{jaffe1976}, and
\citeboth{bodie1976} who both found a negative relationship between inflation and stock returns.
To account for this,
\citeboth{mogdiliani1979} proposed the Inflation Illusion Hypothesis, 
arguing that investors overestimate the discount rate in valuations, leading to mispricing. 
\citeboth{gultekin1983} and \citeboth{firth1979}, however, found that the relationship between nominal stock returns and inflation in the UK is positive, and that real returns on UK stocks have remained stable even as inflation varied.
Later, \citeboth{hasan2008} provided evidence against this, challenging the Fisher hypothesis and enabling 
the use of inflation to

@@

Interest rates, being the cost of borrowing, directly impact firms by increasing financing costs, reducing profitability, and lowering growth prospects. For consumers, higher rates raise borrowing costs, leading to reduced spending and demand, which further lowers corporate profitability. 
The empirical research supports this theory: \citeboth{alam2009} found a significant inverse relationship between interest rates and stock prices across multiple countries from 1988 to 2003; 
\citeboth{talla2013} observed a negative relationship between interest rates and stock prices on the Stockholm Stock Exchange; 
\citeboth{demir2019} reported similar findings for the Turkish Stock Exchange. In the case of the UK, \citeboth{neifar2023} used a Non-Linear ARDL model to find a significantly negative long-run relationship between interest rates and the UK stock market. Higher interest rates also encourage a shift from stocks to safer 
assets like bonds, this is known as the flight-to-quality phenomenon, where rising bond yields make bonds more attractive compared to stocks. \citeboth{asgharian2016} provided evidence for this, particularly in developed markets like the UK. This suggests a typically negative correlation between stock prices and interest rates.

@@

There is a large body of literature regarding time-series forecasting. First, 
a discussion on the feasibility of such forecasting is presented. Following this, machine learning methods for forecasting are 
discussed.

@@

The consistent success of various high-profile investors in the decades 
after the development of the EMH have challenged 
its central argument, something acknowledged in the revised formulation of the EMH 
by \citeboth{fama1991}. Furthermore, the empirical research is not in a
greement on whether the London Stock Exchange is efficient at any of the three levels. 
\citeboth{libberton2010} and \citeboth{rounaghi} give evidence of weak-form efficiency,
however, \citeboth{borges2010}, \citeboth{asghar2023} and \citeboth{bhavsar2015} al provide evidence to reject the EMH. 
Some studies such as \citeboth{rosini2020} even found that the 
efficiency of the FTSE100 and FTSE250 varied over time. The empirical findings then suggest 
that it is not impossible to forecast movements in the 
price of stock indices on the FTSE250 using macroeconomic variables.

@@

Before constructing the ARDL model a natural logarithmic transform is applied
on the variables, this is done because in a log-transformed model,
the coefficients can be interpreted as elasticities. Put simply, 
a coefficient in a model, where both the dependent and independent variables
are in logarithmic form, represents the percentage change in the 
dependent variable for a one percent change in an independent variable, 
which furnishes us with easily interpretable results. Having done this,
the level at which the variables are stationary is checked.
Various techniques exist for stationarity testing, however in this study the 
Augmented Dickey-Fuller (ADF) unit root test, was used.
